---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# concrete
[![CRAN](http://www.r-pkg.org/badges/version/concrete)](http://www.r-pkg.org/pkg/concrete)
[![R-CMD-check](https://github.com/blind-contours/concrete/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/blind-contours/concrete/actions/workflows/R-CMD-check.yaml)
[![Coverage Status](https://coveralls.io/repos/github/blind-contours/concrete/badge.svg?branch=main)](https://coveralls.io/github/blind-contours/concrete?branch=main)

> Continuous-Time Targeted Maximum Likelihood Estimation for Survival Analysis with Competing Risks 

## Description

`concrete` is an R package designed to use targeted maximum likelihood
estimation (TMLE) to compute covariate-adjusted marginal cumulative incidence
estimates in right-censored survival settings with and without competing risks. 

---

## Installation

For standard use, we recommend installing the package from
[CRAN](https://cran.r-project.org/) via

```{r cran-installation, eval = FALSE}
install.packages("concrete")
```

You can install a stable release of `concrete` from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/) with:

```{r gh-installation, eval = FALSE}
devtools::install_github("blind-contours/concrete")
```

---

## Advanced TMLE controls

The TMLE update can be configured through `formatArguments()`. The default
stopping rule is the original relative empirical EIC criterion. For rare-event
settings, a hybrid rule can be more stable because it combines the original
relative threshold with a small absolute tolerance:

```{r tmle-control, eval = FALSE}
ConcreteArgs <- formatArguments(
  DataTable = data,
  EventTime = "time",
  EventType = "status",
  Treatment = "trt",
  Intervention = 0:1,
  TargetTime = c(365, 730),
  TargetEvent = 1,
  UpdateMethod = "adaptive",
  EICStopRule = "hybrid",
  EICStopAbsTol = 1e-3
)
```

Use `getTmleDiagnostics()` on a fitted object to inspect final component-wise
EIC ratios, the update trace, or the norm trajectory:

```{r diagnostics, eval = FALSE}
ConcreteEst <- doConcrete(ConcreteArgs)
getTmleDiagnostics(ConcreteEst, type = "components")
getTmleDiagnostics(ConcreteEst, type = "trace")
```

## Survival learner library

Hazard models can mix Cox formulas with optional survival learners. The
following aliases are supported in `Model[[event_type]]`: `"coxnet"`, `"rsf"`
or `"randomForestSRC"`, `"aareg"` or `"additive_hazards"`, and `"hal"` or
`"hal9001"`.

```{r hazard-learners, eval = FALSE}
Model <- list(
  trt = c("SL.glm", "SL.glmnet"),
  "0" = list(Cox = survival::Surv(time, status == 0) ~ .,
             RSF = "rsf",
             Aareg = "aareg"),
  "1" = list(Cox = survival::Surv(time, status == 1) ~ .,
             HAL = "hal")
)
```

The optional learners require their corresponding packages when selected:
`glmnet` for Coxnet, `randomForestSRC` for random survival forests, and
`hal9001` for HAL.

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/blind-contours/concrete/issues).
