---
title: "Convergence diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convergence diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE
)
```

This article explains how to evaluate the TMLE update when `concrete` reports
slow convergence or non-convergence. The goal is to distinguish a meaningful
targeting problem from a numerically tiny empirical efficient influence curve
(EIC) component in a rare-event setting.

## What convergence means here

The TMLE update tries to make the empirical mean of each target EIC component
small. The original stopping rule checks:

```text
abs(PnEIC) <= seEIC / (sqrt(n) * log(n))
```

This is a relative, component-specific threshold. It can become extremely strict
when an event is rare or a component has near-zero variability.

`concrete` now exposes three stopping rules:

| Rule | Meaning | Typical use |
|---|---|---|
| `relative` | Original component-specific rule | Default, best first check |
| `absolute` | `abs(PnEIC) <= EICStopAbsTol` | Sensitivity analyses on absolute-risk scale |
| `hybrid` | `abs(PnEIC) <= max(relative threshold, EICStopAbsTol)` | Rare events or near-zero EIC variance |

A setting such as `EICStopRule = "hybrid"` and `EICStopAbsTol = 1e-3` means:
keep the original relative criterion unless it is stricter than an absolute
empirical EIC tolerance of 0.001 on the risk scale.

## Inspect the diagnostics

```{r diagnostics}
components <- getTmleDiagnostics(ConcreteEst, type = "components")
components[order(ratio, decreasing = TRUE)]

trace <- getTmleDiagnostics(ConcreteEst, type = "trace")
trace

norm <- getTmleDiagnostics(ConcreteEst, type = "norm")
norm
```

The component table is usually the most useful first view.

Key columns:

- `PnEIC`: empirical mean EIC for the component
- `RelativeCriteria`: original relative stopping threshold
- `AbsoluteCriteria`: absolute threshold supplied by `EICStopAbsTol`
- `StopCriteria`: threshold used by the selected rule
- `ratio`: `abs(PnEIC) / StopCriteria`
- `check`: whether that component passed the selected rule
- `Converged`: whether the overall update converged
- `ConvergenceStep`: update step where convergence was reached

Focus first on rows with `check == FALSE`, sorted by `ratio`.

```{r failing-components}
components[check == FALSE][order(ratio, decreasing = TRUE)]
```

## Recommended escalation ladder

Start simple and add flexibility only after the conservative analysis behaves as
expected.

### 1. Run a conservative baseline

Use default Cox hazards and a small treatment Super Learner library.

```{r conservative-baseline}
Model <- list(
  arm = c("SL.mean", "SL.glm"),
  "0" = list(Censor = survival::Surv(time, event == 0) ~ arm + age + sex),
  "1" = list(Event = survival::Surv(time, event == 1) ~ arm + age + sex)
)

ConcreteArgs <- formatArguments(
  DataTable = trial,
  EventTime = "time",
  EventType = "event",
  Treatment = "arm",
  ID = "id",
  Intervention = makeITT(),
  TargetTime = c(365, 730),
  TargetEvent = 1,
  CVArg = list(V = 5),
  Model = Model,
  Verbose = FALSE
)

ConcreteEst <- doConcrete(ConcreteArgs)
```

### 2. Use adaptive updating

The adaptive method uses a line search with rollback and is the recommended
first convergence fix.

```{r adaptive}
ConcreteArgs$UpdateMethod <- "adaptive"
ConcreteArgs <- formatArguments(ConcreteArgs)
ConcreteEst <- doConcrete(ConcreteArgs)
```

### 3. Use the hybrid stopping rule for rare events

```{r hybrid}
ConcreteArgs$UpdateMethod <- "adaptive"
ConcreteArgs$EICStopRule <- "hybrid"
ConcreteArgs$EICStopAbsTol <- 1e-3
ConcreteArgs <- formatArguments(ConcreteArgs)
ConcreteEst <- doConcrete(ConcreteArgs)
```

Use this when the largest failing components have very small absolute `PnEIC`
values but large ratios because the relative threshold is tiny.

### 4. Increase iterations only if progress is continuing

```{r max-iter}
ConcreteArgs$MaxUpdateIter <- 1000
ConcreteArgs <- formatArguments(ConcreteArgs)
ConcreteEst <- doConcrete(ConcreteArgs)
```

If the trace has flattened and the same tiny components remain, increasing
iterations may not change the practical estimate.

### 5. Simplify or stabilize nuisance estimation

If `abs(PnEIC)` remains large, the issue may be nuisance instability rather than
only the stopping threshold.

Try:

- simpler hazard learner libraries
- fewer high-variance learners for small trials
- stronger propensity-score truncation through `MinNuisance`
- fewer target times for initial debugging
- checking for arms with no events near the target time

## Interpreting common patterns

| Pattern | Likely meaning | Next step |
|---|---|---|
| Large `ratio`, tiny `AbsPnEIC` | Relative rule is too strict on a near-zero component | Try `hybrid` with `1e-3` |
| Large `ratio`, large `AbsPnEIC` | Targeting problem remains meaningful | Use adaptive update and inspect learners |
| Many failing components for censoring | Censoring or positivity instability | Check censoring by arm and covariates |
| Failure only for one rare event/time | Sparse target component | Report event counts and try hybrid rule |
| Norm decreases then rebounds | Update overshooting | Use `UpdateMethod = "adaptive"` |

## A reporting template

For trial reports or issue reports, record:

```{r reporting-template}
list(
  package_version = as.character(packageVersion("concrete")),
  update_method = ConcreteArgs$UpdateMethod,
  eic_stop_rule = ConcreteArgs$EICStopRule,
  eic_stop_abs_tol = ConcreteArgs$EICStopAbsTol,
  max_update_iter = ConcreteArgs$MaxUpdateIter,
  target_time = ConcreteArgs$TargetTime,
  target_event = ConcreteArgs$TargetEvent,
  event_counts = trial[, .N, by = .(arm, event)],
  components = getTmleDiagnostics(ConcreteEst, type = "components")
)
```

This is usually enough to understand whether a convergence issue is numerical,
data-sparsity related, or learner related.
