---
title: "Learner library guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learner library guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE
)
```

This guide explains the learner options exposed through `formatArguments()`.
There are two separate nuisance-learning tasks:

- treatment assignment models, supplied through the `SuperLearner` package
- censoring and event hazard models, supplied as a candidate hazard library

The hazard library currently works as a cross-validated discrete selector: for
each censoring or event type, `concrete` evaluates the candidate hazard learners
and uses the learner with the lowest validation loss.

## Conservative first library

For first use in a trial, start with simple and stable learners.

```{r conservative}
Model <- list(
  arm = c("SL.mean", "SL.glm"),
  "0" = list(Censor = survival::Surv(time, event == 0) ~ arm + age + sex),
  "1" = list(Event = survival::Surv(time, event == 1) ~ arm + age + sex)
)
```

For competing risks, add one hazard model list for each positive event code.

```{r competing}
Model <- list(
  arm = c("SL.mean", "SL.glm"),
  "0" = list(Censor = survival::Surv(time, event == 0) ~ arm + age + sex),
  "1" = list(Death = survival::Surv(time, event == 1) ~ arm + age + sex),
  "2" = list(Competing = survival::Surv(time, event == 2) ~ arm + age + sex)
)
```

## Treatment Super Learner

For randomized trials, the treatment model is often simple. If treatment was
randomized 1:1 and the analysis is intent-to-treat, a simple library is a good
starting point:

```{r treatment-sl-simple}
Model <- list(
  arm = c("SL.mean", "SL.glm")
)
```

For observational or covariate-adaptive settings, add flexible learners already
available through `SuperLearner`.

```{r treatment-sl-flexible}
Model <- list(
  arm = c("SL.mean", "SL.glm", "SL.glmnet", "SL.xgboost")
)
```

Only include treatment learners whose packages are installed and appropriate
for the sample size.

## Hazard learner aliases

Hazard learners are specified inside `Model[["0"]]`, `Model[["1"]]`, and other
event-specific entries.

| Alias | Learner | Package |
|---|---|---|
| Cox formula | Cox proportional hazards | `survival` |
| `"coxnet"` | Penalized Cox model | `glmnet` |
| `"rsf"` or `"randomForestSRC"` | Random survival forest | `randomForestSRC` |
| `"aareg"` or `"additive_hazards"` | Additive hazards | `survival` |
| `"hal"` or `"hal9001"` | HAL pooled discrete-time hazard | `hal9001` |

Optional packages:

```{r optional-installs}
install.packages(c("glmnet", "randomForestSRC", "hal9001"))
```

To test optional hazard learners on a small built-in example before trying your
own data:

```{r optional-smoke}
Sys.setenv(CONCRETE_RUN_OPTIONAL_LEARNERS = "true")
source(system.file("examples", "trialist-smoke-test.R", package = "concrete"))
```

When optional learners are installed, the smoke test prints a summary like this:

| analysis | status | elapsed_sec | converged | step | max_ratio | failing_components |
|---|---|---:|---|---:|---:|---:|
| cox_only | ok | 1.2 | TRUE | 4 | 0.743 | 0 |
| additive_hazards | ok | 1.2 | TRUE | 4 | 0.899 | 0 |
| coxnet | ok | 2.3 | TRUE | 13 | 0.838 | 0 |
| rsf | ok | 1.1 | TRUE | 4 | 0.748 | 0 |
| hal | ok | 1.5 | TRUE | 4 | 0.747 | 0 |

This table is not a benchmark. It is a quick installation and learner-path
check. On real trial data, compare estimates, convergence diagnostics, and
runtime across the learner ladder.

## Cox plus machine-learning hazards

This example gives each event type a small candidate library. The selected
hazard learner can differ across censoring, the event of interest, and competing
events.

```{r hazard-library}
Model <- list(
  arm = c("SL.mean", "SL.glm", "SL.glmnet"),
  "0" = list(
    Cox = survival::Surv(time, event == 0) ~ arm + age + sex + albumin,
    Coxnet = "coxnet",
    Aalen = "aareg"
  ),
  "1" = list(
    Cox = survival::Surv(time, event == 1) ~ arm + age + sex + albumin,
    RSF = "rsf",
    HAL = "hal"
  ),
  "2" = list(
    Cox = survival::Surv(time, event == 2) ~ arm + age + sex + albumin,
    RSF = "rsf"
  )
)

ConcreteArgs <- formatArguments(
  DataTable = trial,
  EventTime = "time",
  EventType = "event",
  Treatment = "arm",
  ID = "id",
  Intervention = makeITT(),
  TargetTime = c(365, 730),
  TargetEvent = 1,
  CVArg = list(V = 5),
  Model = Model,
  UpdateMethod = "adaptive",
  EICStopRule = "hybrid",
  EICStopAbsTol = 1e-3,
  Verbose = FALSE
)

ConcreteEst <- doConcrete(ConcreteArgs)
```

## Suggested libraries by trial size

These are starting points, not rules.

| Setting | Suggested treatment library | Suggested hazard library |
|---|---|---|
| Small trial or rare event | `SL.mean`, `SL.glm` | Cox formulas, additive hazards |
| Moderate trial | `SL.mean`, `SL.glm`, `SL.glmnet` | Cox, Coxnet, additive hazards |
| Larger trial with nonlinear risk | add tree/boosting learners | Cox, Coxnet, RSF, HAL |
| First convergence debugging | `SL.mean`, `SL.glm` | Cox only |

## Inspect selected hazard learners

When `ReturnModels = TRUE`, fitted objects keep initial learner information.

```{r inspect-learners}
fits <- attr(ConcreteEst, "InitFits")

# Treatment model Super Learner weights.
fits[["arm"]]

# Hazard model selection risks are stored on each fitted hazard object.
lapply(fits[setdiff(names(fits), "arm")], function(fit) {
  attr(fit, "HazSL")
})
```

The hazard learner output records cross-validated risks and the selected
candidate for each event type. A simplified example looks like:

```text
$`1`
$`1`$SupLrnCVRisks
      Cox    Coxnet       RSF       HAL
   118.20    116.75    117.40    119.05

$`1`$SLCoef
   Cox Coxnet    RSF    HAL
     0      1      0      0
```

Here, `Coxnet` was selected for event type `1`. The exact object names depend on
the names you used in `Model`.

## Practical testing sequence

For a trial testing week, use the same estimand and data set across a ladder of
learner libraries.

```{r testing-sequence}
model_cox <- list(
  arm = c("SL.mean", "SL.glm"),
  "0" = list(Cox = survival::Surv(time, event == 0) ~ .),
  "1" = list(Cox = survival::Surv(time, event == 1) ~ .)
)

model_penalized <- list(
  arm = c("SL.mean", "SL.glm", "SL.glmnet"),
  "0" = list(Cox = survival::Surv(time, event == 0) ~ ., Coxnet = "coxnet"),
  "1" = list(Cox = survival::Surv(time, event == 1) ~ ., Coxnet = "coxnet")
)

model_flexible <- list(
  arm = c("SL.mean", "SL.glm", "SL.glmnet"),
  "0" = list(Cox = survival::Surv(time, event == 0) ~ ., Coxnet = "coxnet"),
  "1" = list(
    Cox = survival::Surv(time, event == 1) ~ .,
    Coxnet = "coxnet",
    RSF = "rsf",
    Aalen = "aareg",
    HAL = "hal"
  )
)
```

Compare:

- point estimates and confidence intervals
- selected hazard learners
- convergence diagnostics
- runtime
- whether the result is stable to removing the most flexible learners
