---
title: "Trial-design and regulatory toolkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trial-design and regulatory toolkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE
)
```

This article collects the features of `concrete` that go beyond a single
covariate-adjusted absolute risk and that map onto how randomized trials are
designed, analyzed, and reviewed. The framing follows two regulatory documents:

- **FDA (2023), _Adjusting for Covariates in Randomized Clinical Trials for
  Drugs and Biological Products_.** Covariate adjustment is encouraged because it
  can improve precision without changing the estimand in a randomized trial. The
  guidance favors collapsible marginal summaries (risk difference, risk ratio,
  restricted mean) over the hazard ratio, which is non-collapsible and hard to
  interpret under non-proportional hazards.
- **ICH E9(R1), the estimand and sensitivity-analysis addendum.** A trial
  analysis should name its estimand through five attributes and pre-specify how
  intercurrent events are handled, accompanied by sensitivity analyses for the
  assumptions that cannot be checked from the data.

Each section below shows the corresponding `concrete` function. The code is
illustrative; substitute your own trial `data.table`. Throughout, `trial` is a
one-row-per-subject table with an observed `time`, an event-type `event`
(`0` = censored, positive integers for events), a binary `arm`, an `id`, and
baseline covariates.

```{r setup}
library(concrete)
library(data.table)

trial <- as.data.table(your_trial_data)
```

## 1. Name the estimand (ICH E9(R1))

`makeEstimand()` records the analysis target using the five E9(R1) attributes.
It does not change the estimation — it documents the target so it can travel
with the results into a statistical analysis plan and the final report.

```{r make-estimand}
est_def <- makeEstimand(
  treatment  = "as randomized (intent-to-treat)",
  population  = "all randomized participants",
  endpoint    = "cause-specific absolute risk of relapse at 730 days",
  summary     = "risk difference",
  strategy    = "treatment policy"
)
est_def
```

## 2. Handle intercurrent events

An intercurrent event (ICE) — treatment discontinuation, rescue medication, a
competing terminal event — is coded as one of the event-type values in the data.
`applyIntercurrentEvent()` recodes the outcome to match the chosen E9(R1)
strategy and returns a copy of the data carrying the estimand as an attribute:

- **Treatment policy** (default ITT): the ICE is ignored, follow-up continues,
  data returned unchanged.
- **Hypothetical**: the ICE is recoded as censoring (`0`). `concrete`'s
  inverse-probability-of-censoring weighting then targets the risk that would be
  seen had the ICE not occurred, **under the strong, untestable assumption that
  the ICE acts as conditionally independent censoring** — pair it with the
  censoring sensitivity analysis in Section 6.
- **Composite**: the ICE is merged into the event of interest, defining a
  composite endpoint.

```{r apply-ice}
# Suppose event type 3 codes treatment discontinuation (the intercurrent event).
# Hypothetical strategy: estimate the risk had discontinuation not occurred.
trial_hyp <- applyIntercurrentEvent(
  trial, EventTime = "time", EventType = "event",
  Intercurrent = 3, strategy = "hypothetical", intercurrent = "discontinuation"
)

# Composite strategy: count discontinuation as an occurrence of event 1.
trial_comp <- applyIntercurrentEvent(
  trial, EventTime = "time", EventType = "event",
  Intercurrent = 3, strategy = "composite", TargetEvent = 1
)
```

The `"while on treatment"` and `"principal stratum"` strategies are intentionally
not supported: the former is a different (descriptive) estimand that an IPCW
estimator does not target, and the latter needs latent stratification beyond the
current scope.

## 3. Cross-fitting (CV-TMLE) for machine-learning nuisances

When the propensity and hazard nuisances are flexible machine-learning fits, the
plug-in carries an empirical-process (overfitting) term that can invalidate
influence-function inference. Cross-fitting removes it by predicting each
subject's nuisances from folds that did not include them. Turn it on with a
single argument:

```{r cross-fit}
args_cf <- formatArguments(
  DataTable = trial, EventTime = "time", EventType = "event", Treatment = "arm",
  ID = "id", Intervention = makeITT(), TargetTime = c(365, 730), TargetEvent = 1,
  CVArg = list(V = 5),
  CrossFit = TRUE
)
est_cf <- doConcrete(args_cf)
```

Cross-fitting is the continuous-time survival analogue of CV-TMLE / double
machine learning, and is what licenses the use of the random-forest, HAL, and
ensemble hazard learners with honest standard errors.

## 4. Ensemble hazard Super Learner

By default `concrete` performs discrete Super Learner: cross-validation *selects*
the single best hazard learner. `HazEnsemble = TRUE` instead fits a convex
*combination* of the library, with weights chosen to minimize the
cross-validated counting-process loss. This is more robust when no single learner
dominates across the time axis.

```{r ensemble}
args_ens <- formatArguments(
  DataTable = trial, EventTime = "time", EventType = "event", Treatment = "arm",
  ID = "id", Intervention = makeITT(), TargetTime = c(365, 730), TargetEvent = 1,
  CVArg = list(V = 5),
  CrossFit = TRUE, HazEnsemble = TRUE,
  Model = list(
    arm = c("SL.glm", "SL.glmnet"),
    "0" = list(Cox = survival::Surv(time, event == 0) ~ ., RSF = "rsf"),
    "1" = list(Cox = survival::Surv(time, event == 1) ~ ., HAL = "hal")
  )
)
est_ens <- doConcrete(args_ens)
```

## 5. Win ratio, win odds, and net benefit

For hierarchical or composite comparisons, `getWinRatio()` turns the targeted
survival curves into a covariate-adjusted, doubly-robust win ratio, win odds, and
net benefit, each with an influence-function confidence interval. Because it is
built from the targeted curves rather than from raw pairwise comparisons, it
inherits `concrete`'s covariate adjustment and censoring handling. Use a
reasonably dense `TargetTime` grid up to the horizon.

```{r win-ratio}
args_wr <- formatArguments(
  DataTable = trial, EventTime = "time", EventType = "event", Treatment = "arm",
  ID = "id", Intervention = makeITT(),
  TargetTime = c(180, 365, 545, 730), TargetEvent = 1, CVArg = list(V = 5)
)
est_wr <- doConcrete(args_wr)

getWinRatio(est_wr, Horizon = 730, Intervention = c(1, 2))
#> reports P(win), P(loss), P(tie), Win Ratio, Win Odds, and Net Benefit
```

A win ratio above 1 (and a net benefit above 0) favors the active arm; the
confidence interval for the win ratio crossing 1 is the corresponding test of no
difference.

For a **prioritized hierarchy** of competing events, pass `TargetEvent` as an
ordered vector (highest priority first). For example, with cause 1 = death and
cause 2 = hospitalization, `TargetEvent = c(1, 2)` decides each pairwise
comparison on death first and breaks ties on hospitalization (the
Pocock / Finkelstein--Schoenfeld rule). The events must have been targeted in
`doConcrete()`, and the comparison uses each patient's first event (the competing-
risks structure `concrete` models).

```{r win-ratio-hier}
args_h <- formatArguments(
  DataTable = trial, EventTime = "time", EventType = "event", Treatment = "arm",
  ID = "id", Intervention = makeITT(),
  TargetTime = c(180, 365, 545, 730), TargetEvent = c(1, 2), CVArg = list(V = 5)
)
getWinRatio(doConcrete(args_h), Horizon = 730, Intervention = c(1, 2),
            TargetEvent = c(1, 2))   # death > hospitalization
```

## 6. Censoring sensitivity (tipping point)

The hypothetical strategy and the core estimator both assume censoring is
independent given the measured covariates (MAR). `senseCensoring()` stress-tests
that assumption: it imputes an increasing fraction `delta` of censored subjects
as having had the event of interest, re-fits the estimator at each `delta`, and
reports the **tipping point** — the smallest fraction at which a significant
conclusion would be overturned.

```{r sense-censoring}
sc <- senseCensoring(
  args_wr,
  deltas = c(0, 0.05, 0.10, 0.15, 0.20),
  Estimand = "RD", Intervention = c(1, 2)
)
sc                          # estimate / CI / p-value at each delta
attr(sc, "tippingPoint")    # smallest delta that overturns the conclusion
```

A conclusion that survives a large `delta` is robust to departures from
independent censoring; one that tips at a small `delta` should be reported with
caution.

## 7. How much did covariate adjustment buy you?

Covariate adjustment does not change the estimand in a randomized trial, but it
can tighten the confidence intervals. Fit an unadjusted (treatment-only) version
and compare with `getRelativeEfficiency()`, which reports the variance ratio.

```{r relative-efficiency}
unadj_args <- formatArguments(
  DataTable = trial, EventTime = "time", EventType = "event", Treatment = "arm",
  ID = "id", Intervention = makeITT(), TargetTime = c(365, 730), TargetEvent = 1,
  CVArg = list(V = 5),
  Model = list(arm = "SL.mean",
               "0" = list(Cox = survival::Surv(time, event == 0) ~ arm),
               "1" = list(Cox = survival::Surv(time, event == 1) ~ arm))
)
unadj_est <- doConcrete(unadj_args)

getRelativeEfficiency(
  Adjusted   = getOutput(est_cf,    Estimand = "RD", Intervention = c(1, 2)),
  Unadjusted = getOutput(unadj_est, Estimand = "RD", Intervention = c(1, 2))
)
```

A `RelEfficiency` above 1 means adjustment was worth it: a value of 1.25 says the
adjusted analysis has the precision of an unadjusted analysis on 25% more
participants.

## Putting it together: an analysis-plan checklist

A reproducible, E9(R1)-aware analysis with `concrete` typically records:

1. The estimand, via `makeEstimand()`, including the intercurrent-event strategy.
2. The data handling for that strategy, via `applyIntercurrentEvent()`.
3. A cross-fitted (`CrossFit = TRUE`), optionally ensemble (`HazEnsemble = TRUE`)
   fit, so machine-learning nuisances keep honest inference.
4. The primary summary — risk difference / ratio (`getOutput()`), restricted mean
   survival time (`getRMST()` / `targetRMST()`), or win ratio (`getWinRatio()`).
5. Pre-specified sensitivity analyses — `senseCensoring()` for the
   independent-censoring assumption.
6. The precision gain from adjustment, via `getRelativeEfficiency()`.

See the [Trialist quickstart](trialist-quickstart.html) for the core workflow and
[How concrete works](how-concrete-works.html) for the estimator itself.
