---
title: "Testing protocol and current limitations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing protocol and current limitations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  eval = FALSE
)
```

This article gives trial analysts a short protocol for testing `concrete` on
their own data and reporting useful feedback.

## Step 1: Install and run the package smoke test

Install the development version from GitHub:

```{r install}
install.packages("remotes")
remotes::install_github("blind-contours/concrete")
```

Then run the built-in smoke test:

```{r smoke-test}
library(concrete)

source(system.file("examples", "trialist-smoke-test.R", package = "concrete"))
```

The smoke test runs a small competing-risk analysis using a Cox-only hazard
library, prints event counts, prints absolute risks, risk differences, and risk
ratios, and returns a `smoke_summary` table.

To also try optional hazard learners that are installed on your machine:

```{r optional-smoke-test}
Sys.setenv(CONCRETE_RUN_OPTIONAL_LEARNERS = "true")
source(system.file("examples", "trialist-smoke-test.R", package = "concrete"))
```

The optional pass attempts additive hazards, Coxnet, random survival forests,
and HAL when the required packages are installed.

## Step 2: Run the conservative analysis on your trial

Use a simple learner library first. This makes it easier to identify whether any
problem is due to the data structure, the estimand, or a flexible learner.

```{r conservative-trial}
Model <- list(
  arm = c("SL.mean", "SL.glm"),
  "0" = list(Censor = survival::Surv(time, event == 0) ~ arm + age + sex),
  "1" = list(Event = survival::Surv(time, event == 1) ~ arm + age + sex)
)

ConcreteArgs <- formatArguments(
  DataTable = trial,
  EventTime = "time",
  EventType = "event",
  Treatment = "arm",
  ID = "id",
  Intervention = makeITT(),
  TargetTime = c(365, 730),
  TargetEvent = 1,
  CVArg = list(V = 5),
  Model = Model,
  UpdateMethod = "adaptive",
  EICStopRule = "hybrid",
  EICStopAbsTol = 1e-3,
  Verbose = FALSE
)

ConcreteEst <- doConcrete(ConcreteArgs)
ConcreteOut <- getOutput(
  ConcreteEst,
  Estimand = c("Risk", "RD", "RR"),
  Intervention = c(1, 2)
)

ConcreteOut
getTmleDiagnostics(ConcreteEst, type = "components")
```

For competing risks, add one model entry for each positive event code.

## Step 3: Compare against your usual analysis

Record the standard analysis that your trial team would usually report.

Useful comparisons:

- event and censoring counts by randomized arm
- unadjusted cumulative incidence or Kaplan-Meier estimates at the same target
  times
- cause-specific Cox model output, when relevant
- adjusted `concrete` absolute risks
- adjusted `concrete` risk differences and risk ratios
- g-computation plug-in estimates from `getOutput(..., GComp = TRUE)`

The Cox hazard ratio is not the same estimand as the marginal risk ratio from
`concrete`, so use it as context rather than as a direct equality check.

## Step 4: Escalate learners

Use the same data and estimand while changing only the learner library.

```{r learner-ladder}
model_cox <- list(
  arm = c("SL.mean", "SL.glm"),
  "0" = list(Cox = survival::Surv(time, event == 0) ~ .),
  "1" = list(Cox = survival::Surv(time, event == 1) ~ .)
)

model_coxnet <- list(
  arm = c("SL.mean", "SL.glm", "SL.glmnet"),
  "0" = list(Cox = survival::Surv(time, event == 0) ~ ., Coxnet = "coxnet"),
  "1" = list(Cox = survival::Surv(time, event == 1) ~ ., Coxnet = "coxnet")
)

model_flexible <- list(
  arm = c("SL.mean", "SL.glm", "SL.glmnet"),
  "0" = list(Cox = survival::Surv(time, event == 0) ~ ., Aalen = "aareg"),
  "1" = list(
    Cox = survival::Surv(time, event == 1) ~ .,
    Coxnet = "coxnet",
    RSF = "rsf",
    Aalen = "aareg",
    HAL = "hal"
  )
)
```

Compare point estimates, runtime, selected learners, and convergence
diagnostics.

## Current limitations

The current public testing target is intentionally narrow.

Supported:

- one row per participant
- right-censored event time outcome
- optional competing risks
- baseline binary treatment coded `0` and `1`
- static interventions such as everyone assigned `A = 1` versus everyone
  assigned `A = 0`
- baseline covariate adjustment
- target absolute risks, risk differences, and risk ratios at prespecified times

Not currently supported in the main trialist workflow:

- longitudinal treatment regimes
- recurrent events
- delayed entry or left truncation
- multi-arm or continuous treatment without custom intervention work
- post-baseline time-varying covariates
- clustered trial designs requiring special variance handling

Use caution when:

- one arm has very few events by the target time
- censoring is highly imbalanced by arm or covariates
- optional machine-learning learners dominate a small trial
- the flexible learner results differ sharply from Cox-only results
- the TMLE update has large absolute `PnEIC` values after the adaptive update

## What to send back

For useful feedback, include:

- package version from `packageVersion("concrete")`
- `sessionInfo()`
- event and censoring counts by treatment arm
- target event and target time
- exact `Model` list
- `UpdateMethod`, `EICStopRule`, and `EICStopAbsTol`
- `getOutput()` table
- `getTmleDiagnostics(ConcreteEst, type = "components")`
- whether the issue also occurs with a Cox-only learner library

GitHub issue templates are available for convergence issues, learner failures,
and estimand questions.
